---
title: "Assignment 2: Panel Data – Temperature Shocks and Economic Growth"
author: "Advanced Methods for Data Analysis in Public Policy Research FS 26 –-- ETH Zürich"
output: html_document
editor_options:
  markdown:
    wrap: sentence
---

> **Topic**: Temperature Shocks and Economic Growth\
> **Group Name**: *Best Friends Forever and Ever*\
> **Authors**: *Ben Berton, Adam Pederson*\
> **Date**: *DD-MM-YYYY*

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

### Setup

```{r data setup, eval = T, echo = T, results = 'hide', message=F, warning=F}
# import all needed packages here
library(dplyr)

# load the data. Place the data in the same folder as your markdown script.
df <- read.csv('climate_growth_teaching.csv')
```

### 1. Panel Structure

**a)** Is the dataset a balanced panel? Explain how you check.

```{r question 1 a, eval = T, echo = T}

# multiplying the number of countries by the number of years
expected_rows = length(unique(df$country)) * length(unique(df$year))
print(expected_rows)

# getting the total number of actual rows
total_rows = nrow(df)
print(total_rows)
```
No, it is not balanced, as there are 9945 expected rows but only 9655 actual rows.

**b)** Report the number of countries and the average number of years observed per country.

```{r question 1 b, eval = T, echo = T}
# getting the number of countries
number_of_countries = length(unique(df$country))
print(number_of_countries)

# Getting the number of years observed for each country
country_obs <- df %>%
  group_by(country) %>%
  summarise(
    unique_years = n_distinct(year)
  )

# Getting the average number of years per country
mean(country_obs$unique_years)
```
There are on average 49.51282 years observed per country

### 2. Descriptive Statistics

**a)** Report summary statistics for `gdp_growth` and `temp_c` (min, max, mean, sd; and selected percentiles).

```{r question 2 a, eval = T, echo = T}
summary(df$gdp_growth)

summary(df$temp_c)

```
For gdp_growth:
     Min.   1st Qu.    Median      Mean   3rd Qu.      Max. 
-0.204053 -0.094437 -0.031397 -0.049745 -0.002289  0.057988 

For temp_c:
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 -1.383  12.985  17.492  17.919  22.528  39.072 

**b)** Pick one country and plot `temp_c` over time. Briefly describe the pattern.

```{r question 2 b, eval = T, echo = T}
# install.packages("ggplot2")
library(ggplot2)

# filter by country 1
c_1_obs <- df %>% filter(country == "Country 1")
ggplot(data = c_1_obs) +
  geom_line(aes(x = year, y = temp_c), color = "red") +
  labs(
    title = "Temperature Over Time for Country 1",
    x = "Time",
    y = "Temperature (°C)"
  )
```
There is not a strong pattern in the data, although tempuratures do appear to 
fluctuate to more high and low extremes from the 1990s onwards. In general, 
annual average tempuratures for country 1 are around 18 degrees.


### 3. Pooled Regression with Clustering

**a)** Regress `gdp_growth` on `temp_c` and `precip_mm` using pooled OLS. Use standard errors clustered at the country level. Interpret the coefficient on `temp_c`.

```{r question 3 a, eval = T, echo = T}
# install.packages("fixest")
library(fixest)
ols_pooled = feols(gdp_growth ~ temp_c + precip_mm, data = df, cluster = ~country)
summary(ols_pooled)
```
	
             Estimate.      Std. Error.    t value.    Pr(>|t|)
(Intercept)	 0.01698161	    0.01249030.  	 1.35958	  1.7554e-01	
temp_c.bb 	-0.00316898.  	0.00058389. 	-5.42734	  1.6946e-07	***
precip_mm 	-0.00000849	    0.00000674. 	-1.26060	  2.0897e-01

Holding precipitation constant, a one unit increase in Celsius is 
associated with a 0.003 decrease in log GDP per capita growth.

**b)** Explain why clustered standard errors at the country level are appropriate for this specification.\
Clustering standard errors at the country level is important, as without it, there
is autocorrelation in the error terms between years for the same country.
Without it, the standard errors would be underestimated, inflating the 
statistical significance.


### 4. Country Fixed Effects

**a)** Estimate a fixed effects model with country FE:
$$gdp\_growth_{it} = \alpha_i + \beta \, temp\_c_{it} + \gamma \, precip\_mm_{it} + u_{it}$$

```{r question 4 a, eval = T, echo = T}
# add your code here

```

**b)** Compare $\beta$ to the pooled estimate. Why might they differ?\

**c)** In one paragraph, explain what country FE control for.\

### 5. Two-Way Fixed Effects

**a)** Add year fixed effects to the model in (4). Compare the temp coefficient to (4).

```{r question 5 a, eval = T, echo = T}
# add your code here

```

**b)** Explain what year FE absorb in this context.\

### 6. Temperature Shocks and Heterogeneity

Dell et al. emphasize within-country shocks and heterogeneous effects by income.

**a)** Regress `gdp_growth` on `temp_dev` and `precip_dev` with country FE and year FE.

```{r question 6 a, eval = T, echo = T}
# add your code here

```

**b)** Add the interaction term `temp_dev_poor`. Interpret the coefficient on `temp_dev_poor`.

```{r question 6 b, eval = T, echo = T}
# add your code here

```

### 7. Robustness / Influential Units

**a)** Identify one or two potentially influential countries and re-estimate your preferred model excluding them. Do results change meaningfully? Explain briefly.

```{r question 7 a, eval = T, echo = T}
# add your code here

```

### 8. Conclusion

**a)** Summarize your main empirical conclusion in 2--3 sentences.\

**b)** Give one caveat about interpreting panel FE estimates as causal in this setting.\
