---
title: "Assignment 2: Panel Data – Temperature Shocks and Economic Growth"
author: "Advanced Methods for Data Analysis in Public Policy Research FS 26 –-- ETH Zürich"
output: html_document
editor_options:
  markdown:
    wrap: sentence
---

> **Topic**: Temperature Shocks and Economic Growth\
> **Group Name**: *Best Friends Forever and Ever*\
> **Authors**: *Ben Berton, Adam Pederson*\
> **Date**: *DD-MM-YYYY*

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

### Setup

```{r data setup, eval = T, echo = T, results = 'hide', message=F, warning=F}
# import all needed packages here
library(dplyr)

# load the data. Place the data in the same folder as your markdown script.
df <- read.csv('climate_growth_teaching.csv')
```

### 1. Panel Structure

**a)** Is the dataset a balanced panel? Explain how you check.

```{r question 1 a, eval = T, echo = T}

# multiplying the number of countries by the number of years
expected_rows = length(unique(df$country)) * length(unique(df$year))
print(expected_rows)

# getting the total number of actual rows
total_rows = nrow(df)
print(total_rows)
```
No, it is not balanced, as there are 9945 expected rows but only 9655 actual rows.

**b)** Report the number of countries and the average number of years observed per country.

```{r question 1 b, eval = T, echo = T}
# getting the number of countries
number_of_countries = length(unique(df$country))
print(number_of_countries)

# Getting the number of years observed for each country
country_obs <- df %>%
  group_by(country) %>%
  summarise(
    unique_years = n_distinct(year)
  )

# Getting the average number of years per country
mean(country_obs$unique_years)
```
There are on average 49.51282 years observed per country

### 2. Descriptive Statistics

**a)** Report summary statistics for `gdp_growth` and `temp_c` (min, max, mean, sd; and selected percentiles).

```{r question 2 a, eval = T, echo = T}
summary(df$gdp_growth)

summary(df$temp_c)

```
For gdp_growth:
     Min.   1st Qu.    Median      Mean   3rd Qu.      Max. 
-0.204053 -0.094437 -0.031397 -0.049745 -0.002289  0.057988 

For temp_c:
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 -1.383  12.985  17.492  17.919  22.528  39.072 

**b)** Pick one country and plot `temp_c` over time. Briefly describe the pattern.

```{r question 2 b, eval = T, echo = T}
# install.packages("ggplot2")
library(ggplot2)

# filter by country 1
c_1_obs <- df %>% filter(country == "Country 1")
ggplot(data = c_1_obs) +
  geom_line(aes(x = year, y = temp_c), color = "red") +
  labs(
    title = "Temperature Over Time for Country 1",
    x = "Time",
    y = "Temperature (°C)"
  )
```
There is not a strong pattern in the data, although tempuratures do appear to 
fluctuate to more high and low extremes from the 1990s onwards. In general, 
annual average tempuratures for country 1 are around 18 degrees.


### 3. Pooled Regression with Clustering

**a)** Regress `gdp_growth` on `temp_c` and `precip_mm` using pooled OLS. Use standard errors clustered at the country level. Interpret the coefficient on `temp_c`.

```{r question 3 a, eval = T, echo = T}
# install.packages("fixest")
library(fixest)
ols_pooled = feols(gdp_growth ~ temp_c + precip_mm, data = df, cluster = ~country)
summary(ols_pooled)
```
	
             Estimate.      Std. Error.    t value.    Pr(>|t|)
(Intercept)	 0.01698161	    0.01249030.  	 1.35958	  1.7554e-01	
temp_c.bb 	-0.00316898.  	0.00058389. 	-5.42734	  1.6946e-07	***
precip_mm 	-0.00000849	    0.00000674. 	-1.26060	  2.0897e-01

Holding precipitation constant, a one unit increase in Celsius is 
associated with a 0.003 decrease in log GDP per capita growth.

**b)** Explain why clustered standard errors at the country level are appropriate for this specification.\

Clustering standard errors at the country level is important, as without it, there is autocorrelation in the error terms between years for the same country. Without it, the standard errors would be underestimated, inflating the statistical significance.


### 4. Country Fixed Effects

**a)** Estimate a fixed effects model with country FE:
$$gdp\_growth_{it} = \alpha_i + \beta \, temp\_c_{it} + \gamma \, precip\_mm_{it} + u_{it}$$

```{r question 4 a, eval = T, echo = T}
# install.packages("plm")
library(plm)
model_fe <- plm(gdp_growth ~ temp_c + precip_mm, 
                  index=c("country"),
                  data=df)
summary(model_fe)
```
Oneway (individual) effect Within Model

Call:
plm(formula = gdp_growth ~ temp_c + precip_mm, data = df, index = c("country"))

Unbalanced Panel: n = 195, T = 46-51, N = 9655

Residuals:
       Min.     1st Qu.      Median     3rd Qu.        Max. 
-0.04604457 -0.00851605  0.00030635  0.00825444  0.04902558 

Coefficients:
             Estimate  Std. Error  t-value Pr(>|t|)    
temp_c    -3.6004e-03  1.0709e-04 -33.6213   <2e-16 ***
precip_mm  1.9283e-07  7.1826e-07   0.2685   0.7883    
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Total Sum of Squares:    1.672
Residual Sum of Squares: 1.4934
R-Squared:      0.10679
Adj. R-Squared: 0.088283
F-statistic: 565.404 on 2 and 9458 DF, p-value: < 2.22e-16

**b)** Compare $\beta$ to the pooled estimate. Why might they differ?\

The beta for temp_c changed from -0.00317 to -0.00360, and precip_mm changed from negative to slightly positive. The values differ because the pooled OLS model suffers from omitted variable bias, blending the true effect of temperature with the fixed differences between countries.

**c)** In one paragraph, explain what country FE control for.\

Country FE controls for the omitted variable bias caused by the permanent characteristics of countries. Without fixed effects, the pooled model mixes the true effect of year-to-year temperature shocks with these unobserved, permanent characteristics.



### 5. Two-Way Fixed Effects

**a)** Add year fixed effects to the model in (4). Compare the temp coefficient to (4).

```{r question 5 a, eval = T, echo = T}
# install.packages("plm")
library(plm)
model_fe <- plm(gdp_growth ~ temp_c + precip_mm, 
                  index=c("country", "year"),
                  data=df)
summary(model_fe)

```
Oneway (individual) effect Within Model

Call:
plm(formula = gdp_growth ~ temp_c + precip_mm, data = df, index = c("country", 
    "year"))

Unbalanced Panel: n = 195, T = 46-51, N = 9655

Residuals:
       Min.     1st Qu.      Median     3rd Qu.        Max. 
-0.04604457 -0.00851605  0.00030635  0.00825444  0.04902558 

Coefficients:
             Estimate  Std. Error  t-value Pr(>|t|)    
temp_c    -3.6004e-03  1.0709e-04 -33.6213   <2e-16 ***
precip_mm  1.9283e-07  7.1826e-07   0.2685   0.7883    
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Total Sum of Squares:    1.672
Residual Sum of Squares: 1.4934
R-Squared:      0.10679
Adj. R-Squared: 0.088283
F-statistic: 565.404 on 2 and 9458 DF, p-value: < 2.22e-16

The temperature coefficient remained virtually unchanged after adding year fixed effects, suggesting that global, time-specific shocks did not significantly bias the initial estimates.

**b)** Explain what year FE absorb in this context.\

Year fixed effects absorb unobserved events that affect all countries simultaneously in a given year, lik global economic shocks.

### 6. Temperature Shocks and Heterogeneity

Dell et al. emphasize within-country shocks and heterogeneous effects by income.

**a)** Regress `gdp_growth` on `temp_dev` and `precip_dev` with country FE and year FE.

```{r question 6 a, eval = T, echo = T}
library(plm)
model_fe <- plm(gdp_growth ~ temp_dev + precip_dev, 
                  index=c("country", "year"),
                  data=df)
summary(model_fe)
```
Oneway (individual) effect Within Model

Call:
plm(formula = gdp_growth ~ temp_dev + precip_dev, data = df, 
    index = c("country", "year"))

Unbalanced Panel: n = 195, T = 46-51, N = 9655

Residuals:
       Min.     1st Qu.      Median     3rd Qu.        Max. 
-0.04604457 -0.00851605  0.00030635  0.00825444  0.04902558 

Coefficients:
              Estimate  Std. Error  t-value Pr(>|t|)    
temp_dev   -3.6004e-03  1.0709e-04 -33.6213   <2e-16 ***
precip_dev  1.9283e-07  7.1826e-07   0.2685   0.7883    
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Total Sum of Squares:    1.672
Residual Sum of Squares: 1.4934
R-Squared:      0.10679
Adj. R-Squared: 0.088283
F-statistic: 565.404 on 2 and 9458 DF, p-value: < 2.22e-16

**b)** Add the interaction term `temp_dev_poor`. Interpret the coefficient on `temp_dev_poor`.

```{r question 6 b, eval = T, echo = T}
library(plm)
model_fe <- plm(gdp_growth ~ temp_dev * poor + precip_dev, 
                index = c("country", "year"),
                effect = "twoways",
                data = df)
summary(model_fe)

```
Twoways effects Within Model

Call:
plm(formula = gdp_growth ~ temp_dev * poor + precip_dev, data = df, 
    effect = "twoways", index = c("country", "year"))

Unbalanced Panel: n = 195, T = 46-51, N = 9655

Residuals:
      Min.    1st Qu.     Median    3rd Qu.       Max. 
-0.0452401 -0.0081453  0.0001945  0.0080216  0.0411003 

Coefficients:
                 Estimate  Std. Error  t-value  Pr(>|t|)    
temp_dev      -9.6549e-04  1.4614e-04  -6.6067 4.143e-11 ***
precip_dev     1.8557e-07  6.9655e-07   0.2664    0.7899    
temp_dev:poor -5.3127e-03  2.0789e-04 -25.5548 < 2.2e-16 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Total Sum of Squares:    1.6645
Residual Sum of Squares: 1.3914
R-Squared:      0.16406
Adj. R-Squared: 0.14212
F-statistic: 615.422 on 3 and 9407 DF, p-value: < 2.22e-16

The interaction term temp_dev:poor has a significantly larger negative coefficient than the baseline temp_dev. This indicates that the economic impact of temperature shocks depends heavily on a country’s wealth. For rich countries, a one-unit increase in temperature deviation results in a 9.65e-04 decrease in GDP growth. However, poor countries have an additional 5.31e-03 penalty, leading to a total growth decrease of approximately 6.28e-03. This confirms that the GDP growth of poor countries is far more sensitive to temperature deviation than that of wealthy nations.

### 7. Robustness / Influential Units

**a)** Identify one or two potentially influential countries and re-estimate your preferred model excluding them. Do results change meaningfully? Explain briefly.

```{r question 7 a, eval = T, echo = T}
# Finding the country and year with the lowest gdp growth
print(df[which.min(df$gdp_growth), ])

# Creating DataFrame without country 71
df_filtered <- subset(df, country != "Country 71")

# Making model without Country 71
model_fe <- plm(gdp_growth ~ temp_dev * poor + precip_dev, 
                index = c("country", "year"),
                effect = "twoways",
                data = df_filtered)
summary(model_fe)

```
Show in New Window
Twoways effects Within Model

Call:
plm(formula = gdp_growth ~ temp_dev * poor + precip_dev, data = df_filtered, 
    effect = "twoways", index = c("country", "year"))

Unbalanced Panel: n = 194, T = 46-51, N = 9605

Residuals:
       Min.     1st Qu.      Median     3rd Qu.        Max. 
-0.04522623 -0.00814397  0.00018992  0.00805746  0.04118051 

Coefficients:
                 Estimate  Std. Error  t-value  Pr(>|t|)    
temp_dev      -9.6516e-04  1.4623e-04  -6.6004 4.322e-11 ***
precip_dev     2.1409e-07  6.9815e-07   0.3066    0.7591    
temp_dev:poor -5.3040e-03  2.0872e-04 -25.4120 < 2.2e-16 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Total Sum of Squares:    1.6545
Residual Sum of Squares: 1.3858
R-Squared:      0.16242
Adj. R-Squared: 0.1404
F-statistic: 604.894 on 3 and 9358 DF, p-value: < 2.22e-16

I identified the country with the largest loss in GDP growth in one year, Country 71, which had a decrease of ~0.2 in annual log GDP per capita growth in 1989. After removing the country from the model, the results did not meaningfully change, although the coefficient for precip_dev increased from 1.8557e-07 to 2.1409e-07. This is not surprising, as there are 195 countries in the dataset, so removing one was not likely to make a difference unless it was an extremely strong outlier.

### 8. Conclusion

**a)** Summarize your main empirical conclusion in 2--3 sentences.\
Based on the model, we can deduce that when controlling for precipitation and two-way fixed effects for country and year, GDP growth decreases slightly as temperatures rise. Furthermore, the results show that this negative effect is significantly stronger for poor countries than for rich ones, suggesting that developing nations are more vulnerable to temperature shocks.

**b)** Give one caveat about interpreting panel FE estimates as causal in this setting.\
It is difficult to interpret these fixed effects estimates as purely causal because the model may still suffer from time-varying omitted variable bias. While fixed effects control for permanent country traits and simultaneous effects across all countries in a given year, they do not account for unobserved factors that change over time within a specific country.
