---
title: "Assignment 1: Logit / Probit"
author: "Advanced Methods for Data Analysis in Public Policy Research FS 26 –-- ETH Zürich"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

> **Topic**: Income Inequality and Happiness\
> **Group Name**: *Best Friends Forever and Ever*\
> **Authors**: *Ben Berton, Adam Pederson*\
> **Date**: *02-03-2026*

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

### Setup

```{r data setup, eval = T, echo = T, results = 'hide', message=F, warning=F}
# import all needed packages here
library(dplyr)

# load the data. Place the data in the same folder as your markdown script.
df <- read.csv('happiness_inequality.csv')
```

### 1. Descriptives

**a)** Report summary statistics (mean, sd, min, max) for `happy`, `ln_income`, `gini`, `age`, `educ_yrs`.

```{r question 1 a, eval = T, echo = T}
## Happy summary statistics
summary(df$happy)
sd(df$happy)

## ln_income summary statistics
summary(df$ln_income)
sd(df$ln_income)

## gini summary statistics
summary(df$gini)
sd(df$gini)

## age summary statistics
summary(df$age)
sd(df$age)

```

**b)** Report the share of $happy$ in the full sample and by unemployment status.

```{r question 1 b, eval = T, echo = T}
# Share of happy
sum(df$happy == 1)/nrow(df)

# Getting employed and unemployed data frames
unemployed_df <- df %>% filter(unemployed == 1)
employed_df <- df %>% filter(unemployed == 0)

# Share of unemployed that are happy
sum(unemployed_df$happy == 1)/nrow(unemployed_df)

# Share of employed that are happy
sum(employed_df$happy == 1)/nrow(employed_df)

```

**c)** Briefly describe whether individuals in higher-inequality regions report lower happiness.

```{r question 1 c, eval = T, echo = T}
# Calculate correlation between gini coefficient and happiness
cor(df$gini, df$happy)

# Calculate mean happiness in locations were the gini is in the top quartile
top_q_gini <- df %>% filter(gini >= quantile(gini, 0.75))
mean(top_q_gini$happy)

# Calculate mean happiness in locations were the gini is in the low quartile
bottom_q_gini <- df %>% filter(gini <= quantile(gini, 0.25))
mean(bottom_q_gini$happy)

# In the top quartile of income inequality, individuals are 62 percent likley 
# to be happy compared to 64 percent of those in low income inequality

```

### 2. Linear Probability Model

**a)** Estimate the LPM: $$happy_i = \alpha + \beta_1 \ln\_income_i + \beta_2 gini_i + \varepsilon_i$$

```{r question 2 a, eval = T, echo = T}
lpm_model <- lm(happy ~ ln_income + gini, data = df)
summary(lpm_model)
```

**b)** Interpret $\beta_1$ and $\beta_2$ as changes in probability.\

Holding the Gini coefficient constant, a one-unit increase in log-income
is associated with a 5.17 percentage point increase in the probability of 
being happy. Holding log-income constant, a one-unit increase in the gini
coefficient is associated with a 20 percentage point decrease in the
probability of being happy.

**c)** Do any fitted values fall outside $[0,1]$?
If yes, report the share and show an example.

```{r question 2 b, eval = T, echo = T}
# Adding the predicted values to the df
df$pred_happiness <- lpm_model$fitted.values

# Checking amout of values outside [0, 1]
range(df$pred_happiness)
summary(df$pred_happiness)

# I was not able to find any fitted values outside of [0, 1], as the lowest 
# value was 0.427 and the highest was 0.7992
```

**d)** In one paragraph, discuss the limitations of the LPM for binary outcomes.\

The limitations of LPMs for binary outcomes are that it can produce values
outside of 0 and 1 and that it doesn't allow the effect of x on the probability
vary depending on the value of X (binary outcomes are normally s shaped).

### 3. Logit and Probit

**a)** Estimate **logit** and **probit** versions of the model specified in Question 2.

```{r question 3 a, eval = T, echo = T}
lpm_model <- lm(happy ~ ln_income + gini, data = df)
logit_model <- glm(happy ~ ln_income + gini, data = df, family = binomial(link = "logit"))
probit_model <- glm(happy ~ ln_income + gini, data = df, family = binomial(link = "probit"))
summary(lpm_model)
summary(logit_model)
summary(probit_model)

```

**b)** Compare the signs and statistical significance of `ln_income` and `gini` across the LPM, Logit, and Probit models.\

For all three models, the variables are statistically significant with a alpha 
of 0, with the log-income variable having a positive sign and the gini
coefficient having a negative sign for each model.

**c)** Explain why logit/probit coefficients are not directly interpretable as percentage-point changes (unlike LPM coefficients).\

They are not directly interpretable as percentage points, because they show a
one unit change in the z value for the probit model and a one unit change
in the log-odds for the logit model. The results can only be interpreted as
probabilities after applying the Logistic or Standard Normal Cumulative
Distribution functions.

### 4. Adding Controls and Clustering

**a)** Estimate a preferred specification (logit or probit) including the following controls: `age`, `age2`, `female`, `educ_yrs`, `married`, `unemployed`.
Use standard errors clustered at the `region` level.

```{r question 4 a, eval = T, echo = T}
# install.packages("fixest")
library(fixest)

probit_model_adv <- feglm(happy ~ ln_income + gini + age + age2 + female
                        + educ_yrs + married + unemployed,
                        data = df,
                        family = binomial(link = "probit"),
                        cluster = ~region)

summary(probit_model_adv)

```

**b)** Report the estimated effects of `ln_income` and `gini` using **marginal effects**.

```{r question 4 b, eval = T, echo = T}
# install.packages("mfx")
library(mfx)

model_mfx <- probitmfx(happy ~ ln_income + gini + age + age2 + female + 
                       educ_yrs + married + unemployed, 
                       data = df, 
                       clustervar1 = "region")

# Print out marginal effects for each var
model_mfx

# A one unit increase in the log-income is associated with a 4.72 percentage
# point increase in happiness, while a one unit increase in the gini coefficient
# is associated with a 21.59 percentage point decrease in happiness

```

**c)** Compare these results to the bivariate model from Question 3.
How do the estimates change?

```{r question 4 c, eval = T, echo = T}
lpm_model <- lm(happy ~ ln_income + gini, data = df)
summary(lpm_model)
# The results were very close, as ln_income showed a 4.72 percentage point 
# increase in the probability of happiness, compared to 5.17 percentage points
# in the bivariatemodel (Question 3). Similarly, the Gini coefficient in the 
# Probit model showed a 21.59 percentage point decrease in happiness, which is 
# very close to the 21.12 percentage point decrease found in Model 3. 
# Both variables remained highly statistically significant (p<0.01)
```

**d)** Which control variables are most strongly associated with happiness?\

Unemployment and marriage are most strongly associated with happiness, as
being unemployed is associated with a 14.3 percentage point decrease in the
probability of being happy, while being married is associated with a 6.8
percentage point increase.

### 5. Predicted Probabilities for Profiles

**a)** Using your preferred nonlinear model from Question 4, compute predicted happiness probabilities for the following profiles:

-   **A:** Age=30, Female=0, Educ=12, Married=0, Unemployed=0, `ln_income` at the 25th percentile, `gini` at the 25th percentile.
-   **B:** Same as A, but `gini` at the 75th percentile.
-   **C:** Same as A, but `ln_income` at the 75th percentile (keep `gini` at the 25th).

```{r question 5 a, echo = T, eval = T}
# Getting ln_income and gini at the 25th and 75th percentile
inc_25 <- quantile(df$ln_income, 0.25)
inc_75 <- quantile(df$ln_income, 0.75)
gini_25 <- quantile(df$gini, 0.25)
gini_75 <- quantile(df$gini, 0.75)

# Creating frames for the different profiles
prediction_profiles <- data.frame(
  age = 30,
  age2 = 900,
  female = 0,
  educ_yrs = 12,
  married = 0,
  unemployed = 0,
  ln_income = c(inc_25, inc_25, inc_75),
  gini = c(gini_25, gini_75, gini_25)
)

# Making the prediction
predictions <- predict(probit_model_adv, newdata = prediction_profiles, type = "response")

```

**b)** Interpret the differences $(B - A)$ and $(C - A)$.

```{r question 5, echo = T, eval = T}
# B - A
predictions[2] - predictions[1]
# Holding individual demographics and log-income constant, a 30-year-old single
# male in a low-inequality region is 1.24 percentage points more likely to be 
# happy than if he were in a high-inequality region.

# C - A
predictions[3] - predictions[1]
# Holding the individual demographics and the gini coefficient constant, a 30 
# year old single male with a high log income is 4.72 percentage points more
# likely to be happy than with low log income.


```

### 6. Marginal Effect of a 10% Income Increase

**a)** Report the **Average Marginal Effect (AME)** of `ln_income`.

```{r question 6 a, eval = T, echo = T}
library(mfx)

model_mfx <- probitmfx(happy ~ ln_income + gini + age + age2 + female + 
                       educ_yrs + married + unemployed, 
                       data = df, 
                       clustervar1 = "region")

# Getting the AME for ln_income
model_mfx$mfxest["ln_income", ]

```

**b)** Based on the AME, approximate the effect of a 10% income increase ($\Delta \ln(x) \approx 0.10$).

A 10% income increase would result in a 0.47 percentage point increase in the
probability of being happy (0.10 * 0.47).

**c)** Explain (in words) why marginal effects can vary across individuals in non-linear models and why the AME is often reported.\

The AME is reported because non linear models assume a s like curve on the data,
so the AME offers a sample-wide summary of this effect, calculated by averaging
the individual marginal effects across all observations.

### 7. Interaction / Heterogeneity

Alesina et al. discuss that inequality may affect groups differently.

**a)** Estimate a model that includes an interaction term between `gini` and `income_z` (relative income).
Interpret the interaction coefficient in terms of probabilities.

```{r question 7, eval = T, echo = T}
model_interaction <- probitmfx(happy ~ ln_income + gini + gini:income_z + 
                                age + age2 + female + educ_yrs + 
                                married + unemployed, 
                                data = df, 
                                clustervar1 = "region")

# Print out marginal effects for each var
model_interaction

# The interaction term between the gini coefficient and income_z having a value
# of -0.0022 means that for every 1-unit increase in the gini coefficient, 
# the marginal effect of relative income on the probability of being happy
# decreases by 0.22 percentage points. It is important to point out that the
# interaction term is not statistically significant though.

```

### 8. Policy Communication and Limitations

**a)** Give two limitations of interpreting coefficient on inequality causally.\

One limitation of interpreting the coefficient on inequality causally is the 
presence of omitted variable bias. Another limitation is that we do not know 
the exact direction of the relationship; this is known as reverse causality. 
This means we cannot be certain if inequality reduces happiness, or if unhappy
populations create unequal economies.

**b)** Write a short policy-facing interpretation (4–6 sentences) of your main result using predicted probabilities or marginal effects.\

The results show that while higher income generally increases happiness, 
regional inequality can dampen these gains. A 10% raise in income is associated
with a 0.47 percentage point increase in the probability of happiness. However,
the negative interaction term (−0.0022) suggests that this boost is less
effective as inequality rises, though this specific interaction did not reach
standard levels of statistical significance. This implies that economic growth 
alone may not maximize public well-being if wealth gaps continue to widen.