---
title: "Assignment 1: Logit / Probit"
author: "Advanced Methods for Data Analysis in Public Policy Research FS 26 –-- ETH Zürich"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

> **Topic**: Income Inequality and Happiness\
> **Group Name**: *Best Friends Forever and Ever*\
> **Authors**: *Ben Berton, Adam Pederson*\
> **Date**: *02-03-2026*

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

### Setup

```{r data setup, eval = T, echo = T, results = 'hide', message=F, warning=F}
# import all needed packages here
library(dplyr)

# load the data. Place the data in the same folder as your markdown script.
df <- read.csv('happiness_inequality.csv')
```

### 1. Descriptives

**a)** Report summary statistics (mean, sd, min, max) for `happy`, `ln_income`, `gini`, `age`, `educ_yrs`.

```{r question 1 a, eval = T, echo = T}
## Happy summary statistics
summary(df$happy)
sd(df$happy)

## ln_income summary statistics
summary(df$ln_income)
sd(df$ln_income)

## gini summary statistics
summary(df$gini)
sd(df$gini)

## age summary statistics
summary(df$age)
sd(df$age)

```

**b)** Report the share of $happy$ in the full sample and by unemployment status.

```{r question 1 b, eval = T, echo = T}
# Share of happy
sum(df$happy == 1)/nrow(df)

# Getting employed and unemployed data frames
unemployed_df <- df %>% filter(unemployed == 1)
employed_df <- df %>% filter(unemployed == 0)

# Share of unemployed that are happy
sum(unemployed_df$happy == 1)/nrow(unemployed_df)

# Share of employed that are happy
sum(employed_df$happy == 1)/nrow(employed_df)

```

**c)** Briefly describe whether individuals in higher-inequality regions report lower happiness.

```{r question 1 c, eval = T, echo = T}
# Calculate correlation between gini coefficient and happiness
cor(df$gini, df$happy)

# Calculate mean happiness in locations were the gini is in the top quartile
top_q_gini <- df %>% filter(gini >= quantile(gini, 0.75))
mean(top_q_gini$happy)

# Calculate mean happiness in locations were the gini is in the low quartile
bottom_q_gini <- df %>% filter(gini <= quantile(gini, 0.25))
mean(bottom_q_gini$happy)

# In the top quartile of income inequality, individuals are 62 percent likley 
# to be happy compared to 64 percent of those in low income inequality

```

### 2. Linear Probability Model

**a)** Estimate the LPM: $$happy_i = \alpha + \beta_1 \ln\_income_i + \beta_2 gini_i + \varepsilon_i$$

```{r question 2 a, eval = T, echo = T}
lpm_model <- lm(happy ~ ln_income + gini, data = df)
summary(lpm_model)
```

**b)** Interpret $\beta_1$ and $\beta_2$ as changes in probability.\

Holding the Gini coefficient constant, a one-unit increase in log-income
is associated with a 5.17 percentage point increase in the probability of 
being happy. Holding log-income constant, a one-unit increase in the gini
coefficient is associated with a 20 percentage point decrease in the
probability of being happy.

**c)** Do any fitted values fall outside $[0,1]$?
If yes, report the share and show an example.

```{r question 2 b, eval = T, echo = T}
# Adding the predicted values to the df
df$pred_happiness <- lpm_model$fitted.values

# Checking amout of values outside [0, 1]
range(df$pred_happiness)
summary(df$pred_happiness)

# I was not able to find any fitted values outside of [0, 1], as the lowest 
# value was 0.427 and the highest was 0.7992
```

**d)** In one paragraph, discuss the limitations of the LPM for binary outcomes.\

The limitations of LPMs for binary outcomes are that it can produce values
outside of 0 and 1 and that it doesn't allow the effect of x on the probability
vary depending on the value of X (binary outcomes are normally s shaped).

### 3. Logit and Probit

**a)** Estimate **logit** and **probit** versions of the model specified in Question 2.

```{r question 3 a, eval = T, echo = T}
lpm_model <- lm(happy ~ ln_income + gini, data = df)
logit_model <- glm(happy ~ ln_income + gini, data = df, family = binomial(link = "logit"))
probit_model <- glm(happy ~ ln_income + gini, data = df, family = binomial(link = "probit"))
summary(lpm_model)
summary(logit_model)
summary(probit_model)

```

**b)** Compare the signs and statistical significance of `ln_income` and `gini` across the LPM, Logit, and Probit models.\

For all three models, the variables are statistically significant with a alpha 
of 0, with the log-income variable having a positive sign and the gini
coefficient having a negative sign for each model.

**c)** Explain why logit/probit coefficients are not directly interpretable as percentage-point changes (unlike LPM coefficients).\

They are not directly interpretable as percentage points, because they show a
one unit change in the z value for the probit model and a one unit change
in the log-odds for the logit model. The results can only be interpreted as
probabilities after applying the Logistic or Standard Normal Cumulative
Distribution functions.

### 4. Adding Controls and Clustering

**a)** Estimate a preferred specification (logit or probit) including the following controls: `age`, `age2`, `female`, `educ_yrs`, `married`, `unemployed`.
Use standard errors clustered at the `region` level.

```{r question 4 a, eval = T, echo = T}
# install.packages("fixest")
library(fixest)
probit_model_adv <- feglm(happy ~ ln_income + gini + age + age2 + female
                        + educ_yrs + married + unemployed,
                        data = df,
                        family = binomial(link = "probit"),
                        cluster = ~region)

summary(probit_model_adv)

```

**b)** Report the estimated effects of `ln_income` and `gini` using **marginal effects**.

```{r question 4 b, eval = T, echo = T}
predictions_ln_income = predict(probit_model_adv,
                                new_data = data.frame("ln_income" = 0,1),
                                type = "response")

```

**c)** Compare these results to the bivariate model from Question 3.
How do the estimates change?

```{r question 4 c, eval = T, echo = T}
# Add your code here

```

**d)** Which control variables are most strongly associated with happiness?\

### 5. Predicted Probabilities for Profiles

**a)** Using your preferred nonlinear model from Question 4, compute predicted happiness probabilities for the following profiles:

-   **A:** Age=30, Female=0, Educ=12, Married=0, Unemployed=0, `ln_income` at the 25th percentile, `gini` at the 25th percentile.
-   **B:** Same as A, but `gini` at the 75th percentile.
-   **C:** Same as A, but `ln_income` at the 75th percentile (keep `gini` at the 25th).

```{r question 5 a, echo = T, eval = T}
# add your code here
```

**b)** Interpret the differences $(B - A)$ and $(C - A)$.

```{r question 5, echo = T, eval = T}
# add your code here

```

### 6. Marginal Effect of a 10% Income Increase

**a)** Report the **Average Marginal Effect (AME)** of `ln_income`.

```{r question 6 a, eval = T, echo = T}
# add your code here

```

**b)** Based on the AME, approximate the effect of a 10% income increase ($\Delta \ln(x) \approx 0.10$).

**c)** Explain (in words) why marginal effects can vary across individuals in non-linear models and why the AME is often reported.\

### 7. Interaction / Heterogeneity

Alesina et al. discuss that inequality may affect groups differently.

**a)** Estimate a model that includes an interaction term between `gini` and `income_z` (relative income).
Interpret the interaction coefficient in terms of probabilities.

```{r question 7, eval = T, echo = T}
# add your code here

```

### 8. Policy Communication and Limitations

**a)** Give two limitations of interpreting coefficient on inequality causally.\

**b)** Write a short policy-facing interpretation (4–6 sentences) of your main result using predicted probabilities or marginal effects.\
