---
title: "Assignment 3: IV Regression — Family Size and Children's Education"
author: "Advanced Methods for Data Analysis in Public Policy Research FS 26 –-- ETH Zürich"
output: html_document
editor_options:
  markdown:
    wrap: sentence
---

> **Topic**: Family Size and Children's Education\
> **Group Name**: *Best Friends Forever and Ever*\
> **Authors**: * Ben Berton, Adam Pedersen*\
> **Date**: *15-04-2026*

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

### Setup

```{r data setup, eval = T, echo = T, results = 'hide', message=F, warning=F}
# import all needed packages here
library(dplyr)
library(haven)

# load the data. Place the data in the same folder as your markdown script.
df <- read_dta('family_size.dta')
```

### 1. Descriptive Statistics

**a)** Summarize `educ_yrs` (min, max, mean, sd) and comment on the distribution.

```{r question 1 a, eval = T, echo = T}
summary(df$educ_yrs)
sd(df$educ_yrs)

```

**Answer:** The distribution of education years is slightly right-skewed, with a mean of 9.44 being higher than the median of 9.00 and a standard deviation of 1.68. Most observations are concentrated between 8 and 11 years, indicating that while the majority of the sample has a basic-to-mid-level education, a small group of high-attaining individuals is pulling the average up.


**b)** What share of children are in families with `morethan2=1`?

```{r question 1 b, eval = T, echo = T}
sum(df$morethan2 == 1) / length(df$morethan2)

```
**Answer:** 55.1% of children have mothers with 3+ children.

### 2. OLS

**a)** Regress `educ_yrs` on `morethan2` using OLS. Interpret the coefficient.

```{r question 2 a, eval = T, echo = T}
model = lm(educ_yrs ~ morethan2, data = df)
summary(model)

```

**Answer:** On average, children whose mothers have three or more children have 0.13 fewer years of education than children from smaller families.
 
 
**b)** Explain why the OLS estimate in (2) may not identify the causal effect of family size on education.\
**Answer:** The OLS estimate in (2) may not identify the causal effect, as family size is an endogenous variable with respect to education. There are unobserved variables, like socioeconomic status, that influence both the expected years of education and the expected number of children for a mother. 

### 3. Instrument Relevance

**a)** Summarize `samesex12`.

```{r question 4 a, eval = T, echo = T}
summary(df$samesex12)

```

**b)** Estimate the first stage: regress `morethan2` on `samesex12`. How large is the effect? Is it statistically significant?

```{r question 4 b, eval = T, echo = T}
inst_model = lm(morethan2 ~ samesex12, data=df)
summary(inst_model)

```

**Answer:** On average, a mother's first two children having the same sex raises the likelihood of having one or more additional children by 7.5 percentage points. It is highly statistically significant with a p < 0.001.

**c)** Report and interpret a weak-instrument diagnostic.

```{r question 4 c, eval = T, echo = T}
summary(inst_model)

```

F-statistic: 1.723e+04 on 1 and 3000184 DF,  p-value: < 2.2e-16
**Answer:** The F-statistic is 17,230, so the instrument is strong.

### 4. Instrument Validity

**a)** Discuss whether `samesex12` is a valid instrument.\

**Answer:** 'samesex12' is likely a valid instrument as it satisfies the two main requirements. First, instrument relevance (corr(Zi,Xi) ≠ 0) is confirmed by the first-stage regression, where the F-statistic of 17,230 indicates a strong relationship. Second, the exclusion restriction corr(Zi,ui) = 0) is supported by the biological randomness of a child's sex at birth. This suggests the instrument is uncorrelated with unobserved factors like mother’s ambition or socioeconomic status, meaning it should only affect the child's education through its impact on family size.


### 5. TSLS / 2SLS Estimation

**a)** Estimate the causal effect of `morethan2` on `educ_yrs` using TSLS with `samesex12` as the instrument. Interpret the estimate.

```{r question 6 a, eval = T, echo = T}
#install.packages("AER")
library(AER)

iv_model <- ivreg(educ_yrs ~ morethan2 | samesex12, data = df)

summary(iv_model)

```

**Answer:** On average, a child's mother having 3+ children lowers their expected years of education by 0.18 years.

### 6. Controls and Precision

**a)** Re-estimate TSLS adding the controls listed above. Do results change?

```{r question 7 a, eval = T, echo = T}
iv_model_ctrls <- ivreg(educ_yrs ~ morethan2 + child_age + child_sex + 
                          mom_age + mom_educ + birth_order + race3 + hispan + year | 
                          samesex12 + child_age + child_sex + 
                          mom_age + mom_educ + birth_order + race3 + hispan + year, 
                        data = df)

summary(iv_model_ctrls)
```

**Answer:** The coefficient changed significantly, shrinking from -0.18 to -0.037, and it is no longer statistically significant (p=0.129). This tells us that in the previous model, the instrument was likely picking up explanatory power from variables that were stuck in the error term. Specifically, factors like mother’s education and child’s age were originally part of the unobserved noise.


**b)** Why might controls help (or not) in IV?\
**Answer:** Controls help by soaking up "seen" noise (like mother’s education), which makes the model more precise and shrinks the standard errors. They also ensure the instrument is valid by accounting for factors that might prevent the instrument from being truly random.

However, they might not help if they are too similar to the instrument. If a control "steals" the instrument's explanatory power, it creates a weak instrument (low F-statistic), making the causal estimate unreliable.

### 7. Balance / Exogeneity Checks

**a)** Can we test the exclusion restriction directly? Explain.\
**Answer:** No, we cannot test the exclusion restriction directly, as the error captured by the instrument is unseen, meaning there is no way to mathematically prove there is no correlation.

**b)** Run at least two balance regressions of predetermined covariates (e.g., mother education, mother age, race/ethnicity) on `samesex12` and interpret.

```{r question 8 b, eval = T, echo = T}
balance_educ <- lm(mom_educ ~ samesex12, data = df)
summary(balance_educ)

balance_age <- lm(mom_age ~ samesex12, data = df)
summary(balance_age)

```

**Answer:** Both the mother’s education and the mother’s age are significantly correlated to the instrument. This means that the instrument is not completely exogenous, as it correlates with the error term. While statistically significant, the magnitudes are so small that the covariates are practically balanced.

### 8. LATE Interpretation

**a)** Interpret your IV estimate as a Local Average Treatment Effect (LATE). Who are the **compliers** in this setting?\
**Answer:** The IV estimate identifies the Local Average Treatment Effect (LATE). This is the causal effect of having a third child specifically for the compliers. The compliers are parents who had a third child only because their first two children were of the same sex (and would not have otherwise).
